---
title: "Using dplyr for data handling"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
---

For these exercises, we will use a new dataset related to mammalian sleep as provided in Data Analysis for the Life Sciences (Rafael Irizarry, Mike Love). Download the CSV file from this location: [https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv).
 
 We are going to read in this data, then test your knowledge of they key `dplyr` functions. We are also going to review two different *classes*: data frames and vectors.
 
###  0 - Load necessary libraries
 0. Load the dplyr library or the tidyverse collection of libraries that includes dplyr.
```{r}
# to be filled
```

###  1 - Read in and inspect the data 
1a. Read in the `msleep_ggplot2.csv` file with the function `read.csv` into an object called `msleep`.
```{r}
# to be filled
```

1b. Use the function `class` to determine what type of object is `msleep`.
```{r}
# to be filled
```

1c. Use the `head` function or `View()` to take a glimpse at the data.
```{r}
# to be filled
```

1d. How many animals are in the data in total and how many are there for each order? Hint: Use `nrow` and `table`
```{r}
# to be filled
```

###  2 - Modify the data
2. Add a new column to the data that contains the fraction of REM sleep of the total sleep using the `mutate` function.
```{r}
# to be filled
```

###  3 - Subset the data to primates
3a. Now use the `filter` function to select only the primates from the `msleep` object and give the resulting object a new name.
```{r}
# to be filled
```

3b. How many primates species are in the data?
```{r}
# to be filled
```

3c. What is the class of the object you obtain after subsetting the table to only include primates?
```{r}
# to be filled
```

###  4 - Calculate the average sleep of primates
4a.  Now use the `select` function to extract the sleep (total) for the primates. 
```{r}
# to be filled
```

4b. What class is this object? 
```{r}
# to be filled
```

4c. Calculate the average amount of sleep for primates using the `mean` function.
```{r}
# to be filled
```

###  5- Advanced exercises
To calculate the average amount of sleep for primates, we could also use the dplyr `summarize` function. We have not introduced this function. For unkown functions it is useful to look at the R help providing details on a function.

5a. Use `?summarize` to open the help page for the `summarize` function. 

5b. Calculate the average amount of sleep for primates starting from the `msleep` object in a single code line using `filter`, the pipe `%>%` and `summarize` to get the answer.
```{r}
# to be filled
```

The `summarize` function can be combined with the `group_by` function to calculate a summary statistics (e.g., the mean) for different groups of rows in a data frame. For example, this allows us to calculate the average amount of sleep for all orders of mammals in the data without having to repeat the exercises 3-6 for all orders separately. 

5c. Take a look at the help of `?group_by`.

5d. Group the data frame obtained in exercise 1 (`msleep`) by "order" and use `summarise` on the grouped data frame to calculate the average amount of sleep for each order.
```{r}
# to be filled
```

In addition to the mean we can compute other summaries, for example the function `n()` counts the number of rows in each group. 

5e. Use `group_by` to group the data frame obtained in exercise 1 by "order" and `summarise` the grouped data frame to generate a new data frame with the average total sleep and number of species per order.
```{r}
# to be filled
```

5f. Use the data frame obtained in exercise 5e and the function `arrange` to sort the rows of the data.frame by the average amount of sleep. Which mammalian order sleeps the most?
```{r}
# to be filled
```

### Final words
6. Call the function `sessionInfo()` to have an output of your R and package versions and operating system. This is good practice to always include so that other people can reproduce your code.
```{r}
# to be filled
```

7. Knit your file to produce a html report on your analysis
