---
title: Welcome!
author: "Britta Velten"
date: "`r Sys.Date()`"
output: html_document
---

# Welcome to the DSCB Data Analysis Course

The purpose of this file is to verify that you have successfully installed R and RStudio.

If you are completely new to R or think a recap would be helpful, I would recommend to watch this [youtube video](https://www.youtube.com/watch?v=p8F_wreHTzw) and go through the Swirl tutorial on *R Programming: The basics of programming in R* (as explained at the end of the video) before our course. In addition, the following resources might be useful:

* Chapter 1-3 of Rafael A. Irizarry's [*Introduction to Data Science*](http://rafalab.dfci.harvard.edu/dsbook/introduction.html)
* [Babraham R Course](https://www.bioinformatics.babraham.ac.uk/training/Introduction_to_R/Introduction%20to%20R%20course%20booklet.pdf)

<mark>**Your task**: Please generate a html version of this file (following the steps described below) and *hand this in via Mattermost*.</mark> 

Make sure you get help if you encounter any problems via the Mattermost channel.


# First steps with R using RStudio and RMarkdown

## Step 0: Getting to know RMarkdown

This is an RMarkdown `.Rmd` file. RMarkdown allows you to combine text with code chunks. Code chunks look like this:

```{r}
# This is a code chunk
# The hash key at the beginning indicates that this line is a comment and not evaluated as code
```

```{r my_code_chunk}
# Code chunks can also have names, here 'my_code_chunk'. 
# These will not appear in the output but can for example be useful to identify chunks giving errors
```
For details on RMarkdown see for example <http://rmarkdown.rstudio.com> and <https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf> or a short video at <https://rmarkdown.rstudio.com/lesson-1.html>


## Step 1: Knit this RMarkdown document to produce an html report

By clicking on the "**Knit**" button on top of this panel we will produce a `.html` report that displays the text and evaluates the code in the `.Rmd` file displaying both the code as well as its output. 

The `.html` file will open automatically and is also saved in the same folder (and with the same name) as the `.Rmd` script.

<mark>**Task 1: Knit this file and verify that you can find and open the `.html` file. Go back to the Rmd file, change the author name at the top of the file to your own name and knit the file again to receive a `.html` file with your name on top.**</mark>


## Step 2: Checking your R version

```{r}
if(version$major < 4 | version$minor < 5) {
  print("Your R version is not up-to-date. Please download the latest R version (>=4.5)")
} else {
  print("Your R version is up-to-date.")
  }
```

<mark>**Task 2 Take a look at the output of this code chunk in the html file to see if your R version is up-to-date.**</mark> If your R version is not up-to-date, you can update it by installing the latest version from this [LINK](https://cran.r-project.org/). After the installation you need to re-open RStudio and re-run this file to check whether you now have the latest version.

## Step 3: Showing the details of your current R session
At the end of each Rmd file it is useful to print the output of the `sessionInfo()` function. This is important for reproducibility as it helps you and others to go back to a html report, which you did in the past and see which operating system, R version and package versions you used.

```{r}
sessionInfo()
```

