---
title: "Using linear regression to predict heights"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
---

We will use the tidyverse and UsingR packge. Load these libraries here (and install them if needed).
```{r}
# to be filled
```

1. The UsingR package contains data on father's height and his son's height. Use `?father.son` to find our more and `head(father.son)` to take a first glimpse at the data.
```{r}
# to be filled
```

2. Draw a scatterplot of the height of father versus son. What is the correlation of the two variables? Would you expect a very different result when using Spearmans' correlation? Verify this.
```{r}
# to be filled
```

3. Now we aim to describe this relationship by a line. To learn the intercept and slope of the data we can use the `lm` function.
```{r}
# to be filled
```

4. Call the `summary` function on the result of the last task. How do you interpret the results?
```{r}
# to be filled
```

4. Let's add the fitted line to our plot using `geom_line` with the learnt slope and intercept.
```{r}
# to be filled
```


5. We can now make a prediction for a son of a 75in-tall father. Create a new data.frame containing the father height for which you want to make predictions. Note the column name must match the ones from the original data frame (father.son). Then use the `predict` function with the new data and the output of the `lm` call to make a prediction of the son's height. Add this point to the plot above.
```{r}
# to be filled
```

6. (optional) Use `plot` on the object created by the `lm` function. How can you use the output to check if your model is an appropriate model?
```{r}
# to be filled
```

```{r}
sessionInfo()
```

