---
title: "Clustering on the CLL expression data"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
---

# Preparations
```{r}
library(tidyverse)
```

# Load and prepare the data
Load the expression data from the MOFAdata package and subset to the patient samples with complete measurements.
```{r}
library(MOFAdata)
data("CLL_data")
data("CLL_covariates")

# obtain RNA data for patients with complete RNA measurements
mRNA <- CLL_data$mRNA # get mRNA data matrix from the CLL data list
mRNA <- t(mRNA) # transpose matrix to have samples in rows
has_no_mRNA <- apply(mRNA, 1, function(x) !any(is.na(x))) # find patient samples with complete mRNA information
complete_patients <- names(has_no_mRNA)[has_no_mRNA]
mRNA_complete <- mRNA[complete_patients,]

# obtain information on Gender and IGHV status for those patients and collect them in a dataframe
IGHV <- CLL_data$Mutations["IGHV", complete_patients]
gender <- CLL_covariates[complete_patients, "Gender"]
df_patinfo <- data.frame(patient_sample = complete_patients, 
                         IGHV = IGHV,
                         gender = gender)
```


# Cluster patients based on expression data

## k-means
1. Use k-means to cluster the patients into two groups based on their normalized expression values.
```{r}
# to be filled
```

2. How do the two clusters compare to the gender and IGHV status?

2a. Use the `table` function with the vector of cluster labels and IGHV to compare.
```{r}
# to be filled
```

2b. Use the `table` function with the vector of cluster labels and gender to compare.
```{r}
# to be filled
```


## Hierarchical clustering
3. Let's use hierarchical clustering to group the patients into two groups.

3a. Calculate the Euclidean distances between patients in the gene expression space using `dist` on the matrix containing mRNA measurements.
```{r}
# to be filled
```

3b. Use the `hclust` function on the distances object (created by `dist`) to obtain the clustering of patients.
```{r}
# to be filled
```

3c. Use `plot` on the output of 1b. to visualize the dendrogram.
```{r}
# to be filled
```

3d. Use the `cut` function on the output of 3b to create a vector containing clusters labels for the patients for 2 clusters.
```{r}
# to be filled
```

3e. Use the `table` command like in ex. 2a and 2b to compare the clusters to the groups defined by IGHV status and gender.
```{r}
# to be filled
```

## Heatmap
4. A heatmap shows the expression values and sorts rows and columns using hierarchical clustering.

4a. Plot a heatmap of the gene expression values starting from the `mRNA_complete` object using `pheatmap`.
```{r}
# to be filled
```

Optional (advanced): 
4b. Annotate your heatmap by IGHV status and gender and the clusters found in exercises 2 and 3.
```{r}
# to be filled
```

4c. Vary the distance measure using `clustering_distance_rows` to use correlation distance instead of `euclidean`. When would this be useful?
```{r}
# to be filled
```

4d. Calculate a correlation matrix specifying the pairwise correlations of all patients. Use `pheatmap` to plot a heatmap of this correlation matrix and again annotate the patients by IGHV status, gender and clusters.
```{r}
# to be filled
```

```{r}
sessionInfo()
```

