---
title: "PCA on the CLL expression data"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
---

# Preparations

0a. Load the following package.
```{r}
library(tidyverse)
```

0b. Install and load the ggfortify package.
```{r}
# to be filled
```

# Load and subset the data

1a. Load the expression data from the MOFAdata package and subset to the patient samples with complete measurements by running the following code and inspect its output.
```{r}
library(MOFAdata)
data("CLL_data")
data("CLL_covariates")

# obtain RNA data for patients with complete RNA measurements
mRNA <- CLL_data$mRNA # get mRNA data matrix from the CLL data list
mRNA <- t(mRNA) # transpose matrix to have samples in rows
has_no_mRNA <- apply(mRNA, 1, function(x) !any(is.na(x))) # find patient samples with complete mRNA information
complete_patients <- names(has_no_mRNA)[has_no_mRNA]
mRNA_complete <- mRNA[complete_patients,]

# obtain information on Gender and IGHV status for those patients and collect them in a dataframe
IGHV <- CLL_data$Mutations["IGHV", complete_patients]
trisomy12 <- CLL_data$Mutations["trisomy12", complete_patients]
gender <- CLL_covariates[complete_patients, "Gender"]
df_patinfo <- data.frame(patient_sample = complete_patients, 
                         IGHV = as.factor(IGHV),
                         gender = gender,
                         trisomy12 = as.factor(trisomy12))
```

1b. How many samples and features are in the `mRNA_complete` object?

# PCA
2 Let's calculate the PCA for the gene expression measurments.
2a. Use the `prcomp` function on the `mRNA_complete` object to calculate a PCA.
```{r}
# to be filled
```

2b. Use the `autoplot` function on your results to plot the PCA for component 1 versus 2 as well as 3 versus 4. Use the `coord_fixed()` option for all plots (what does it do?). 
```{r}
# to be filled
```

2c. Color the plots from 2b first by Gender, then by IGHV and then by trisomy 12.  
```{r}
# to be filled
```

2d. Look at the `rotation` matrix in the PCA object to find which genes contribute most to the first prinicipal component.
```{r}
# to be filled
```

2e. Use the `summary` function on the PCA object to get information on the variance explained by each component
```{r}
# to be filled
```

2f. Extract the variance explained and make an Elbow plot to determine a good number of PCs to keep.
```{r}
# to be filled
```

# Visualize clusters in the PCA plot
3a. Cluster the patients into 3 groups using kmeans.
3b. Make a PCA plot of the first two components colouring patients by the clusters.
```{r}
# to be filled
```

# Non-linear dimension reduction methods (optional, advanced)
4. Plot a tSNE or a UMAP of the patients' expression data using `Rtsne` or `umap`.
```{r}
# to be filled
```

# PCA on other omic layers
5. Optional: Compare the structure between patient samples that you found in the PCA plot from gene expression data to the structure in the drug response data.
5a. How does a PCA of the patient on the drug response data looks like? 
5b. Are the clusters found on the expression data still visible? 
5c. How does IGHV or tisomy12 relate to it?
```{r}
# to be filled
```

```{r}
sessionInfo()
```

